---
title: "Assignment 2"
author: "Lukas Unruh, Teije Langelaan, Gidon Quint"
date: "`r format(Sys.Date(), '%d %B, %Y')`"
output: 
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(333)
library(lme4)
library(Matrix)
library(carData)
library(ggplot2)
options(digits=3)
```

# Exercise 1

```{r}
fruitfly_data=read.table(file="fruitflies.txt",header=TRUE)
```

## Lukas)

### a)

### b)

### c)

### d)

### e)

## Gid)
```{r}
fruitfly_data=read.table(file="fruitflies.txt",header=TRUE)
```
### a)
```{r}
head(fruitfly_data)
fruitfly_data$loglongevity = log(fruitfly_data$longevity)
head(fruitfly_data)

par(mfrow = c(2, 3))

for (activity_level in unique(fruitfly_data$activity)) {
  data_subset = subset(fruitfly_data, activity == activity_level)$loglongevity
  
  qqnorm(data_subset, main = paste("Q-Q plot -", activity_level))
  qqline(data_subset)
  
  hist(data_subset, main = paste("Histogram -", activity_level), xlab = "Log(Longevity)", breaks = 20)
  
  shapiro_test = shapiro.test(data_subset)
  cat("Shapiro-Wilk test for", activity_level, "activity: p-value =", shapiro_test$p.value, "\n")
}

boxplot(loglongevity ~ activity, data = fruitfly_data, main = "Log(Longevity) by Activity Level", xlab = "Activity Level", ylab = "Log(Longevity)")

kruskal.test(loglongevity ~ activity, data = fruitfly_data)

aggregate(loglongevity ~ activity, data = fruitfly_data, FUN = median)

```
The aggregate function's output shows the median log-longevity for each activity group: isolated flies have the highest median log-longevity (4.13), followed by those with low sexual activity (4.03), and flies with high sexual activity have the lowest median log-longevity (3.69). This suggests that increased sexual activity may be associated with decreased longevity in fruit flies

### b)
I do linear regression although we would need normality. However because we use the loglongevity the use of linear regression is fine

```{r}
multireg_model = lm(loglongevity ~ activity + thorax, data = fruitfly_data)
summary(multireg_model)

qqnorm(residuals(multireg_model))
qqline(residuals(multireg_model))
plot(fitted(multireg_model), residuals(multireg_model))
abline(h = 0)
```
This shows that the residuals deviade from normality. Especially at the tails. 
The plot of fitted values against residuals does not show a clear pattern, which is good as it suggests homoscedasticity (constant variance of the residuals across the range of fitted values). There is no apparent funnel shape or systematic pattern, which is indicative of potential issues with variance.

because the Q-Q plot does show some deviation from normality, it could be beneficial to either transform the data to achieve normality or use a robust regression method that is less sensitive to non-normality.

activityisolated and activitylow are both positive, with isolated having a larger coefficient than low. Showing that compared to high sexual activity, being isolated is associated with an increase in log-longevity (which translates to an increase in actual longevity), while low sexual activity is also associated with an increase in log-longevity but to a lesser extent than isolation. Therefore, sexual activity appears to decrease longevity compared to no activity, with higher activity decreasing it more.

```{r}

average_thorax = mean(fruitfly_data$thorax)

new_data = expand.grid(activity = unique(fruitfly_data$activity), thorax = average_thorax)

new_data$predicted_loglongevity = predict(multireg_model, newdata = new_data)

new_data$predicted_longevity = exp(new_data$predicted_loglongevity)

print(new_data)


```

### c)

```{r}
ggplot(fruitfly_data, aes(x = thorax, y = loglongevity, color = activity)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Relationship between Thorax Length and Log(Longevity) by Activity Group",
       x = "Thorax Length",
       y = "Log(Longevity)")

# added interaction between thorax and activity for regression model
interaction_model = lm(loglongevity ~ thorax * activity, data = fruitfly_data)
summary(interaction_model)


qqnorm(residuals(interaction_model))
qqline(residuals(interaction_model))

plot(fitted(interaction_model), residuals(interaction_model))
abline(h = 0)

```
The scatter plot with regression lines shows a positive relationship between thorax length and log longevity. 
The slopes of the lines appear to differ between groups, suggesting that the rate of increase in longevity with thorax length may vary by sexual activity level.

The regression analysis suggests that thorax length positively affects longevity in fruit flies, with the effect being less for flies with higher sexual activity.
The interaction between thorax length and isolated sexual activity is marginally (0.55) significant, indicating a possible difference in this effect between flies with no sexual activity and those with high sexual activity.

GPT GENERATED

The regression analysis conducted so far has shown that there is a significant interaction between thorax length and the isolated group of fruit flies, suggesting that the dependence of longevity on thorax length differs when flies are kept in isolation compared to those with high sexual activity. For flies with low sexual activity, however, the interaction was not statistically significant, indicating that their longevity dependence on thorax length does not differ much from flies with high sexual activity. This suggests that the influence of thorax length on longevity is not consistent across all conditions of sexual activity.


### d)

Model c) has a slightly higher adjusted R-squared value (0.709 compared to 0.717) and these additional explanatory variables are significant (as seen in the interaction term with p = 0.055), this suggests that model c) is potentially better at explaining the variability in the response, despite being more complex.
Another point worth mentioning is that thorax length allows for a more nuanced interpretation of the data, acknowledging the biological reality that physical traits can affect longevity.
However neither is necessarily "wrong".

### e)

```{r}

# this plot as it is done in the lecture but there are probably better ways to do it
plot(longevity ~ thorax, data = fruitfly_data, pch = as.character(fruitfly_data$activity),
     main = "Longevity vs Thorax Length by Activity Level",
     xlab = "Thorax Length", ylab = "Longevity")



fruitfly_data$activity = as.factor(fruitfly_data$activity)

ancova_model = lm(longevity ~ activity + thorax, data = fruitfly_data)

summary(ancova_model)

anova(ancova_model)

drop1(ancova_model, test = "F")



```
The ANCOVA with raw longevity data has a slightly lower adjusted R-squared compared to the model with log-transformed longevity, which suggests that the log transformation may provide a better fit. The choice of using the log of longevity as the response variable likely helped to stabilize variance and linearize the relationship between predictors and longevity, making it a wise choice for the analysis.

## TJ)

### a)

### b)

### c)

### d)

### e)

# Exercise 2

```{r}
birthweight_data=read.csv(file="Birthweight.csv",header=TRUE)

```
In our data set we have multiple continuous and discrete variables. However, for the model we do not have to state that discrete variables are factors as all of them only have 2 levels (no = 0 and yes = 1).
## Lukas)

### a)

### b)

### c)

### d)

### e)

### f)

### g)

### h)

### i)

## Gid)

### a)

### b)

### c)

### d)

### e)

### f)

### g)

### h)

### i)

## TJ)
### a)

```{r}
# Assuming birthweight_data is your dataset
summary(birthweight_data)

# Fit a linear model
TJ_BW <- lm(Birthweight ~ Length + Headcirc + Gestation + mage + mnocig + mheight + mppwt + fage + fedyrs + fnocig + fheight, data=birthweight_data)

# Plot diagnostics for influential points
par(mfrow=c(2,2))
plot(TJ_BW, which=1:4) # Plots for residuals, leverage, etc.
round(cooks.distance(TJ_BW),2)


```

Upon inspection of Cook's distances, we conclude that there are no influence points in the data set, as no observation has a Cook's distance greater than 1.

```{r}
predictors <- birthweight_data[, c("Length", "Headcirc", "Gestation", "mage", "mnocig", "mheight", "mppwt", "fage", "fedyrs", "fnocig", "fheight")]
print(cor(predictors))
library(carData)
vif_values <- vif(TJ_BW)
print(vif_values)
```

Upon inspection of correlation matrix, we note that variables mage and fage appear to be highly correlated (0.806), however their VIF values are under 5, therefore we conclude that the model has no problem of collinearity.

### b)
```{r}
summary(TJ_BW)
TJ_BW_reduced <- step(TJ_BW, direction = "backward")
summary(TJ_BW_reduced)
```
Using the step-down approach we arrived at the following model: Birthweight ~ 
Length + Headcirc + Gestation + mage + mppwt + fheight with R-squared: 0.757

```{r}
plot(TJ_BW_reduced, which = 1)

qqnorm(residuals(TJ_BW_reduced))
qqline(residuals(TJ_BW_reduced))
shapiro.test(residuals(TJ_BW_reduced))
```
In the residuals vs fitted plot, the pattern from 3.0 to 3.5 could hint at non-linearity, adding polynomial or interaction terms might improve fit. Upon further inspection of the QQ-plot and shapiro-Wilk test we conclude that the assumption of normally distributed residuals is not met.

### c)
```{r}
# Calculate the average values for the predictors
avg_data <- data.frame(
  Length = mean(birthweight_data$Length, na.rm = TRUE),
  Headcirc = mean(birthweight_data$Headcirc, na.rm = TRUE),
  Gestation = mean(birthweight_data$Gestation, na.rm = TRUE),
  mage = mean(birthweight_data$mage, na.rm = TRUE),
  mppwt = mean(birthweight_data$mppwt, na.rm = TRUE),
  fheight = mean(birthweight_data$fheight, na.rm = TRUE)
)

# Assuming your model is named 'TJ_BW_reduced' and is already fitted with the correct formula
# Predict the Birthweight for the average values with 95% confidence interval
confidence_interval <- predict(TJ_BW_reduced, newdata = avg_data, interval = "confidence", level = 0.95)

# Predict the Birthweight for the average values with 95% prediction interval
prediction_interval <- predict(TJ_BW_reduced, newdata = avg_data, interval = "prediction", level = 0.95)

# View the results
confidence_interval
prediction_interval
```

### d)
```{r}
library(glmnet)
set.seed(3)
x=as.matrix(birthweight_data[,-3]) #remove the response variable
y=as.double(as.matrix(birthweight_data[,3])) #only the response variable
train=sample(1:nrow(x),0.67*nrow(x)) # train by using 2/3 of the data
x.train=x[train,]; y.train=y[train] # data to train
x.test=x[-train,]; y.test=y[-train] # data to test the prediction quality
lasso.mod=glmnet(x.train,y.train,alpha=1)
cv.lasso=cv.glmnet(x.train,y.train,alpha=1,type.measure="mse")
plot(lasso.mod,label=T,xvar="lambda") #have a look at the lasso path
plot(cv.lasso) # the best lambda by cross-validation
lambda.min=cv.lasso$lambda.min; lambda.1se=cv.lasso$lambda.1se
coef(lasso.mod,s=cv.lasso$lambda.1se) #beta’s for the best lambda
y.pred=predict(lasso.mod,s=lambda.1se,newx=x.test) #predict for test
mse.lasso=mean((y.test-y.pred)^2) #mse for the predicted test rows
```
Lasso model: Birthweight ~ Length + Headcirc + Gestation + lowbwt
Step down:   Birthweight ~ Length + Headcirc + Gestation + mage + mppwt + fheight

The Lasso model and the step-down method both identify Length, Head Circumference, and Gestation as significant predictors of Birthweight, indicating their critical role. The Lasso model is more parsemonious, using two fewer variables. It uniquely includes lowbwt, a binary variable distinguishing between birth weights below and above 6 pounds, which the step-down model omits. Conversely, the step-down model retains mage, mppwt, and fheight, suggesting these factors may have some predictive value not captured by the Lasso model, likely due to its penalty on model complexity that excludes less impactful variables.

### e)

### f)

### g)

### h)

### i)

# Exercise 3

```{r}
awards_data=read.table(file="awards.txt",header=TRUE)
```

## Lukas)

### a)

### b)

### c)

## Gid)

### a)

### b)

### c)

## TJ)

### a)

### b)

### c)
