---
title: "Assignment 2 - Group 9"
author: "Lukas Unruh, Teije Langelaan, Gidon Quint"
date: "`r format(Sys.Date(), '%d %B, %Y')`"
output: 
  pdf_document:
    latex_engine: xelatex
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 3)
set.seed(333)
library(lme4)
library(Matrix)
library(carData)
library(car)
library(ggplot2)
library(glmnet)

options(digits=3)
```

*We used the library lme4, glmnet, Matrix, car and ggplot2 with
options(digits=3) and a seed of 333*

# Exercise 1

```{r}
fruitfly_data=read.table(file="fruitflies.txt",header=TRUE)
fruitfly_data$loglongevity <- log(fruitfly_data$longevity)
```

## a)

```{r}
boxplot(loglongevity ~ activity, data = fruitfly_data, 
        xlab = "Activity Group", ylab = "Log Longevity",
        main = "Log Longevity by Activity Group")
```

We perform ANOVA to test whether sexual activity influences longevity.
ANOVA does assume normality, however, our TA mentioned that we do not
need to check for normality.

-   $H_0$: The mean loglongevity is the same across all activity groups.

-   $H_1$: The mean loglongevity is not the same across all activity
    groups.

```{r}
longaovA <- lm(loglongevity ~ activity, data = fruitfly_data); anova(longaovA)
```

Since we get a p-value of 1.8e-7, we reject $H_0$ and conclude that
sexual activity influences longevity.

The estimated longevities for the three conditions are (asked the TA and
we assume normality, therefore checking the means):

```{r}
tapply(fruitfly_data$longevity, fruitfly_data$activity, mean)
```

From the values you can see that the mean of "high" is a lot lower than
the mean of the two other groups. This suggests that increased sexual
activity may be associated with decreased longevity in fruit flies.

## b)

-   $H_0$: The mean loglongevity is the same across all activity groups
    after accounting for the effect of thorax.

-   $H_1$: The mean loglongevity is not the same across all activity
    groups after accounting for the effect of thorax.

```{r}
longaovB <- lm(loglongevity ~ thorax + activity, data = fruitfly_data)
anova(longaovB)
```

Since we get a p-value of 4e-9, we reject $H_0$ and conclude that sexual
activity influences longevity.

```{r}
summary(longaovB)
```

Activityisolated and activitylow are both positive, with isolated having a larger coefficient than low. Showing that compared to high sexual activity, being isolated is associated with an increase in log-longevity (which translates to an increase in actual longevity), while low sexual activity is also associated with an increase in log-longevity but to a lesser extent than isolation. Therefore, sexual activity appears to decrease longevity compared to no activity, with higher activity decreasing it more.

```{r}
average_thorax = mean(fruitfly_data$thorax)
new_data = expand.grid(activity = unique(fruitfly_data$activity), thorax = average_thorax)
new_data$predicted_loglongevity = predict(longaovB, newdata = new_data)
new_data$predicted_longevity = exp(new_data$predicted_loglongevity)
print(new_data)
```

## c)

```{r, echo=FALSE}
# Define colors for each activity level
activity_colors <- setNames(c("green", "blue", "red"), c("isolated", "low", "high"))

# Base plot with points. No points will be plotted here because 'pch' is set to NA.
plot(loglongevity ~ thorax, data = fruitfly_data, pch = NA,
     main = "Log(Longevity) by Thorax Length and Activity Level",
     xlab = "Thorax Length", ylab = "Log(Longevity)")

# Loop through each activity level to plot points
for (activity_level in names(activity_colors)) {
  # Subset the data for the current activity level
  data_subset <- subset(fruitfly_data, activity == activity_level)
  points(data_subset$thorax, data_subset$loglongevity, col = activity_colors[activity_level], pch = 16)
}

# Loop through each activity level and draw regression lines with specified color
for (activity_level in names(activity_colors)) {
  # Subset the data for the current activity level
  data_subset <- subset(fruitfly_data, activity == activity_level)
  # Check if the subset is not empty to avoid the error
  if(nrow(data_subset) > 0) {
    # Fit the linear model
    fit <- lm(loglongevity ~ thorax, data = data_subset)
    # Generate new data for plotting the line
    new_thorax <- seq(min(data_subset$thorax), max(data_subset$thorax), length.out = 100)
    preds <- predict(fit, newdata = data.frame(thorax = new_thorax))
    # Draw the line
    lines(new_thorax, preds, col = activity_colors[activity_level], lwd = 2)
  }
}

# Add a legend to the plot to identify the colors
legend("bottomright", legend = names(activity_colors), 
       col = activity_colors, pch = 16, lty = 1, cex = 0.8,
       title = "Activity Level")
```

The scatter plot with regression lines shows a positive relationship between thorax length and loglongevity, but no clear interaction clear difference in this effect between sexual activity groups.

We perform ANCOVA:

-   $H_0$: There is no interaction effect between activity and thorax on
    loglongevity.

-   $H_1$: There is an interaction effect between activity and thorax on
    loglongevity.

```{r}
interaction_model = lm(loglongevity ~ thorax * activity, data = fruitfly_data)
anova(interaction_model)
summary(interaction_model)
```

The interaction between thorax length and sexual activity is non significant p-value of 0.15. The interaction between thorax length and isolated sexual activity is marginally (p-value = 0.055) significant, indicating a possible difference in this effect between flies with no sexual activity and those with high sexual activity

## d)
```{r}
model_summary <- summary(longaovA) 
adjusted_r_squared <- model_summary$adj.r.squared
cat("R-adjusted model A: ", adjusted_r_squared)
model_summary <- summary(longaovB) 
adjusted_r_squared <- model_summary$adj.r.squared
cat("\nR-adjusted model B: ", adjusted_r_squared)
```
Model A without thorax interaction has a R-squared = 0.332, model B with thorax interaction has a R-squard = 0.709. Therefore, we should prefer model B. Additionally, model B is model A with an added significant variable namely thorax. However, none of the analyses are wrong.

## e)

-   $H_0$: There is no difference in the mean longevity among different
    activity levels after accounting for thorax length.
-   $H_1$: There is a difference in the mean longevity among different
    activity levels after accounting for thorax length.

```{r}
fruitfly_data$activity = as.factor(fruitfly_data$activity)
ancova_model = lm(longevity ~ thorax + activity, data = fruitfly_data)
summary(ancova_model)
anova(ancova_model)
drop1(ancova_model, test = "F")
```

```{r, echo=FALSE}
#plot heteroskedacity
par(mfrow = c(1, 2))  # Plot side by side
qqnorm(residuals(longaovB), main = "Log Longevity Residuals")
qqline(residuals(longaovB))
qqnorm(residuals(ancova_model), main = "Longevity Residuals")
qqline(residuals(ancova_model))
```

```{r}
shapB <- shapiro.test(residuals(longaovB))
shapanc <-shapiro.test(residuals(ancova_model))
cat("p-value Log Longevity", shapB$p.value)
cat("\np-value Longevity", shapanc$p.value)
```

```{r, echo=FALSE}
par(mfrow = c(1, 2))
plot(longaovB, which = 1, main = "log longevity: Residuals vs Fitted")
plot(ancova_model, which = 1, main = "longevity: Residuals vs Fitted")
plot(residuals(longaovB),fruitfly_data$loglongevity)
plot(residuals(ancova_model), fruitfly_data$longevity)
```



The ANCOVA model using raw longevity data shows a slightly lower adjusted R-squared than the model with log-transformed longevity, hinting that logarithmic transformation may offer a tighter fit. However, this transformation shifts the sample's distribution away from normality, evident from QQ-plots and a Shapiro-Wilk test p-value of 0.058. Before concluding, it's noteworthy that the log transformation appears to marginally equalize the variance of residuals. Despite this, the deviation from normal distribution in the log-transformed model leads us to prefer the raw longevity model. We believe it is unwise to convert variables before concluding the original variable is problematic.

# Exercise 2

```{r}
birthweight_data=read.csv(file="Birthweight.csv",header=TRUE)
```

## a)

```{r, echo=FALSE}
TJ_BW <- lm(Birthweight ~ Length + Headcirc + Gestation + mage + mnocig + mheight + mppwt + fage + fedyrs + fnocig + fheight, data=birthweight_data)
par(mfrow=c(1,3))
plot(TJ_BW, which=1:2)
plot(TJ_BW, which=4)
```

Upon inspection of plots and Cook's distances, we conclude that there are no influence points in the data set, as no observation has a Cook's distance greater than 1.

```{r}
vif_values <- vif(TJ_BW)
print(vif_values)
cor_fage_mage <- cor(birthweight_data$mage, birthweight_data$fage)
print(cor_fage_mage)
```

Upon inspection of correlation matrix, we note that variables mage and fage appear to be highly correlated (0.807), however their VIF values are under 5, therefore we conclude that the model has no problem of collinearity.

## b)

```{r}
final_model = lm(Birthweight ~ Headcirc + Gestation, data=birthweight_data)
summary(final_model)
```

Order of removal: fage, mheight, fedyrs, fnocig, mnocig, Length,
fheight, mage, mppwt


-   $H_0$: The residuals from the linear regression model follow a
    normal distribution.
-   $H_1$: The residuals from the linear regression model do not follow
    a normal distribution.

```{r}
residuals = residuals(final_model)
test_results_sw = shapiro.test(residuals)
cat("The p-value of the Shapiro-Wilk test is:", test_results_sw$p.value)
```

```{r, echo=FALSE, fig.height=3}
par(mfrow = c(1, 3))
boxplot(residuals, main = "Boxplot of Residuals", ylab = "Residuals", xlab = "Index")
qqnorm(residuals, main = "Q-Q Plot of Residuals")
qqline(residuals, col = "red")
hist(residuals, main = "Histogram of Residuals", xlab = "Residuals", breaks = 20)
```

We do not reject $H_0$ so we cannot conclude that the residuals are not normally distribtuted, however after taking a good look at the QQ plot and the histoplot we conclude that our normality of residual distribution is questionable.

## c)

```{r}
fitted_values = fitted(final_model)
average_values = data.frame(
  Headcirc = mean(birthweight_data$Headcirc, na.rm = TRUE),
  Gestation = mean(birthweight_data$Gestation, na.rm = TRUE)
)
confidence_intervals = predict(final_model, newdata = average_values, interval =
                                 "confidence", level = 0.95)
prediction_intervals = predict(final_model, newdata = average_values, interval = 
                                 "prediction", level = 0.95)
print("95% Confidence Intervals for the Mean Response:")
print(confidence_intervals)
print("95% Prediction Intervals for a New Observation:")
print(prediction_intervals)
```

The output indicates that for babies with average head circumference and gestation, the model predicts a mean Birthweight of 3.31 kg, with a 95% confidence interval of 3.21 to 3.42 kg for the mean and a wider 95% prediction interval of 2.61 to 4.02 kg for individual observations, reflecting the increased uncertainty in predicting specific outcomes.

## d)

```{r}
par(mfrow = c(1,2))
x=as.matrix(birthweight_data[,-3]) # remove the response variable
y=as.double(as.matrix(birthweight_data[,3])) # only the response variable
train=sample(1:nrow(x),0.67*nrow(x)) # train by using 2/3 of the data
x.train=x[train,]; y.train=y[train] # data to train
x.test=x[-train,]; y.test=y[-train] # data to test the prediction quality
lasso.mod=glmnet(x.train,y.train,alpha=1)
cv.lasso=cv.glmnet(x.train,y.train,alpha=1,type.measure='mse')
plot(lasso.mod,label=T,xvar="lambda") # have a look at the lasso path
plot(cv.lasso) # the best lambda by cross-validation
lambda.min=cv.lasso$lambda.min; lambda.1se=cv.lasso$lambda.1se
coef(lasso.mod,s=cv.lasso$lambda.1se) # beta’s for the best lambda
y.pred=predict(lasso.mod,s=lambda.1se,newx=x.test) # predict for test
mse.lasso=mean((y.test-y.pred)^2) # mse for the predicted test rows
new_model = lm(Birthweight ~ Length + Headcirc + Gestation + lowbwt, data=birthweight_data)
summary(new_model)
```

Upon inspection of the summary tables, although the LASSO model has a slightly higher R-squared value, we prefer the step-down model, because the two added variables in the LASSO model are insignificant and it does not make sense to add them in respect to complexity.

## e)

Effect of whether mothers smokes on low birth weight

-   $H_0$: Smoking status of the mother has no effect on the likelihood
    of the baby being born with low birth weight.

-   $H_1$: Babies born to smoking mothers are more likely to have low
    birth weight compared to those born to non-smoking mothers.

Effect of mother's age over 35 on low birth weight

-   $H_0$: There is no difference in the likelihood of having a baby
    with low birthweight between mothers over the age of 35 and those 35
    or younger.

-   $H_1$: Babies born to mothers over the age of 35 are more likely to
    have low birthweight compared to those born to younger mothers.

```{r, echo=FALSE}
par(mfrow = c(1, 4))

# Boxplot for Birthweight by Smoking Status
boxplot(birthweight_data$Birthweight ~ birthweight_data$smoker, 
        main = "bw by smoking", 
        xlab = "smoker", ylab = "birthweight", 
        names = c("no", "yes"))

# Boxplot for Birthweight by Mother's Age
boxplot(birthweight_data$Birthweight ~ birthweight_data$mage35, 
        main = "bw by mage34", 
        xlab = "mother aged 35+", ylab = "birthweight", 
        names = c("no", "yes"))

# Barplot for Low Birthweight by Smoking Status
smoker_lowbwt <- table(birthweight_data$lowbwt, birthweight_data$smoker)
barplot(smoker_lowbwt, beside = TRUE, 
        main = "lowbwt by smoking", 
        xlab = "smoker", ylab = "count", 
        names.arg = c("no", "yes"), 
        legend.text = c("no", "yes"), 
        args.legend = list(title = "lowbwt", x = "topright"))

# Barplot for Low Birthweight by Mother's Age
mage35_lowbwt <- table(birthweight_data$lowbwt, birthweight_data$mage35)
barplot(mage35_lowbwt, beside = TRUE, 
        main = "lowbw by mage35", 
        xlab = "mother aged 35+", ylab = "count", 
        names.arg = c("no", "yes"), 
        legend.text = c("no", "yes"), 
        args.legend = list(title = "lowbwt", x = "topright"))
```

```{r} 
table(birthweight_data$smoker, birthweight_data$lowbwt) 
table(birthweight_data$mage35, birthweight_data$lowbwt)
```

The boxplot do not show a clear difference in birthweight resulting from
the varibales 'smoker' and 'mage35'. Both the bar graph, and contingency
table do suggests a potential influence of smoking on low birthweight.
The limited sample sizes \<5 for some experimental units, hinder us in
getting reliable chi-squared results on these possible effects.

## f)

-   $H_0$: The length of gestation (in weeks) has no effect on the
    likelihood of the baby being born with low birthweight.
-   $H_1$: Shorter gestation periods are associated with a higher
    likelihood of the baby being born with low birthweight.

```{r}
TJ_LOWBWT <- glm(lowbwt ~ Gestation + mage35 + smoker, data = birthweight_data, family = binomial)
drop1(TJ_LOWBWT, test="Chisq")
TJ_LOWBWT2 <- glm(lowbwt ~ Gestation + smoker, data = birthweight_data, family = binomial)
drop1(TJ_LOWBWT2, test="Chisq")
exp(coef(TJ_LOWBWT2))
```

In analyzing factors affecting low birthweight (lowbwt), 'mage35' was found not to significantly influence lowbwt (p-value = 0.941), suggesting maternal age over 35 doesn't notably alter the odds of low birthweight. For 'Gestation', each additional week reduces the odds of lowbwt by approximately 77% (odds ratio = 0.23, p-value = 9e-06). 'Smoker' status increases the odds of lowbwt by 241 times (odds ratio = 241, p-value = 0.013), pointing to a substantial risk increase due to smoking. These results highlight the importance of gestation duration and the detrimental impact of smoking on birthweight.

## g)
Interaction gestation and smoking
-   $H_0$: There is no interaction effect between gestation length and
    smoking status on the likelihood of low birthweight.

-   $H_1$: There is an interaction effect between gestation length and
    smoking status on the likelihood of low birthweight.

Interaction gestation and mage35
-   $H_0$: There is no interaction effect between gestation length and
    maternal age over 35 on the likelihood of low birthweight

-   $H_1$: There is an interaction effect between gestation length and
    maternal age over 35 on the likelihood of low birthweight

```{r, warning=FALSE}
model_gestation_smoker = glm(lowbwt ~ Gestation * smoker, data = birthweight_data, family = binomial)
anova_gestation_smoker <- anova(model_gestation_smoker, test="Chisq")
p_value_gestation_smoker <- anova_gestation_smoker["Gestation:smoker", "Pr(>Chi)"]
model_gestation_mage35 = glm(lowbwt ~ Gestation * mage35, data = birthweight_data, family = binomial)
anova_gestation_mage35 <- anova(model_gestation_mage35, test="Chisq")
p_value_gestation_mage35 <- anova_gestation_mage35["Gestation:mage35", "Pr(>Chi)"]
print(paste("P-value for Gestation:smoker interaction:", p_value_gestation_smoker))
print(paste("P-value for Gestation:mage35 interaction:", p_value_gestation_mage35))
```

Both interaction terms—'Gestation:smoker' and 'Gestation:mage35'—do not demonstrate statistically significant effects on low birthweight. The p-value for 'Gestation:smoker' is 0.198, and for 'Gestation:mage35', it is 0.42. These p-values indicate that the interactions between gestation length and either smoking status or maternal age over 35 do not significantly alter the risk of low birthweight in these models. We therefore choose the model without interaction terms.

## h)

```{r}
model_g <- glm(lowbwt ~ Gestation + smoker, data = birthweight_data, family="binomial")
new_data <- data.frame(Gestation = c(40, 40),
                       smoker = c(0, 1))
probabilities <- predict(model_g, newdata = new_data, type = "response")
print(probabilities)
```

For babies born at 40 weeks of gestation, the predicted probability of low birth weight is \<0.001% for non-smokers and 0.02% for smokers, highlighting the significant impact of maternal smoking on birth weight outcomes. This underscores the importance of smoking cessation interventions during pregnancy to mitigate the risk of low birth weight.

## i)

```{r}
contingency_table <- table(birthweight_data$lowbwt, birthweight_data$smoker)
chi_squared_result <- chisq.test(contingency_table)
print(chi_squared_result)
```

Applying a contingency table test, such as the chi-squared test, to address the questions is a valid approach with certain limitations. A major advantage of this method is its straightforwardness and its capacity to explore associations between categorical variables without assuming a specific relationship form. However, one significant disadvantage is its unreliability with very small sample sizes (e.g., \<5 in some groups), which can make chi-squared results misleading, and is this case for this dataset. Additionally, while chi-squared tests are simple and useful for initial explorations, they lack the ability to adjust for other variables or to interpret effects in terms of odds ratios, unlike logistic regression. This simplicity might lead to overlooking complex nuances in the data.

# Exercise 3

```{r}
awards_data=read.table(file="awards.txt",header=TRUE)
```

## a)

-   $H_0$: The type of program (vocational, general, academic) does not
    influence the number of awards earned by students.
-   $H_1$: The type of program (vocational, general, academic) has an
    influence on the number of awards earned by students.

```{r}
awards_data$prog = factor(awards_data$prog)
model = glm(num_awards ~ prog, family = poisson, data = awards_data)
summary(model)
new_data = data.frame(prog = factor(c(1, 2, 3)))
predicted_awards = predict(model, newdata = new_data, type = "response")
print(predicted_awards)
```


Significant p-values for prog2 (general program) and the intercept in the Poisson regression model suggest that program type affects student awards. Students in the general program are likely to receive more awards than those in vocational ones, shown by significant coefficients. Although academic program students also tend to get more awards, it's not statistically significant. Thus, the general program is the most favorable for receiving awards.


## b)

-   $H_0$: There is no difference in the distribution of the number of
    awards across the different program types (vocational, general,
    academic).
-   $H_1$: At least one program type has a significantly different
    distribution of the number of awards compared to the others.

```{r}
kruskal_test_result = kruskal.test(num_awards ~ prog, data = awards_data)
cat("The p-value of the Shapiro-Wilk test is:", kruskal_test_result$p.value)
```

We reject H0 that everything is the same as the p-value is 0.005. However we can not infer which program type is the best for the number of awards.

## c)

```{r}
awardsinteractionmath = glm(num_awards ~ factor(prog)*math, family="poisson", data=awards_data)
summary(awardsinteractionmath)
awardsmath = glm(num_awards ~ factor(prog)+math, family="poisson", data=awards_data)
summary(awardsmath)
new_data = data.frame(prog = factor(c(1, 2, 3), levels = c(1, 2, 3)),
                       math = 56)
predictions = predict(awardsmath, newdata=new_data, type="response")
new_data$predicted_awards = predictions
print(new_data)
```

No significant interaction results were found therefor we removed the interaction. Looking at the additive model with math included, we conclude that program 3 is the best with a predicted amount of awards of 1.213.
